Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru

### 2020-08-10 text/ecography/round_3/enso_comp.Rmd
migrate gitlab repo to github

title: "Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru"
bibliography: transect_refs.bib
csl: ecography.csl
# biblio-style: "natbib"
link-citations: true
---

```{r setup, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("here")
# make sure that figures are allowed to float (hold the assigned position and are not always put on top of a page)
knitr::opts_chunk$set(fig.pos = 'H')
```

@@

Studying the relationship between environmental gradients and species turnover has become a cornerstone of biogeography and community ecology since Humbold's and Bonpland's travels to South America more than 200 years ago [@macarthur_geographical_1984; @whittaker_gradient_1967]. 
Naturally, water availability drives floristic composition and biomass production in the (semi-)arid regions such as NW Peru [@muenchow_coupling_2013; @muenchow_woody_2013].
However, water availability does not only vary in space but also in time, especially in regions heavily affected by the El Niño Southern Oscillation (ENSO).
<!--
A super-humid El Niño (EN) episode benefits other species than a super-dry and fire-bringing La Niña (LN) episode [@richter_monitoring_2005].
-->

ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
El Niño represents the warm and La Niña the cold phase of this cycle. 
ENSO neutral refers to periods when neither El Niño nor La Niña are present. 
La Niña usually causes the reverse effects of El Niño. 
The effects depend on the region and include among others excessive rainfalls, droughts as well as temperature, wind and monsoon anomalies. 
<!--
The anomalies show recurrent patterns to a certain degree but are still highly variable especially on a local scale. 
For instance, in 2016 the predicted Super-Niño rainfalls largely failed to materialize along the coasts of Ecuador and Peru.
Just a year later, unexpected, torrential rainfalls hit the same coasts during a supposedly ENSO neutral year with a magnitude comparable to that of the Super-Niño event in 1997/98. Originating from a so-called coastal EN, the 2017 rainfalls came as a complete surprise, and caused the unprepared population unbearable hardship [@fraser_peru_2017; @ramirez_understanding_2017].
-->

Though extreme climate events like El Niño and La Niña change and reshape ecosystems drastically [@Kogan.2017], ecological research mostly focusses on the effects of El Niño.
Nevertheless, the eco-climatic impact of La Niña is comparable in magnitude with that of El Niño.
For instance, La Niña episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
In any case, ENSO certainly drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño may vary from region to region [@phillips_changes_2009].
Overall, the effect of  extreme events is even pronounced in (semi)-arid ecosystems [@holmgren_extreme_2006], which is especially true for ENSO's terrestrial core region in NW Peru and SW Ecuador [@muenchow_woody_2013; @rollenbeck_climatic_2015] -- one of the most variable regions on Earth in terms of climate [@wehrden_global_2012].
This region is also home to the highly tropical dry forest (TDF) ecosystems.
TDFs are characterized by a high rate of endemism and feature an outstanding biodiversity which is almost as high as that of humid tropical forests [@espinosa_what_2011;@pennington_tropical_2018]. 
At the same time, they are the most threatened tropical ecosystem [@miles_global_2006]. 
In Peru, 95% of TDFs have already disappeared [@portillo-quintero_extent_2010]. 
Even so, tropical dry ecosystems are chronically understudied [@escribano-avila_biodiversity_2017;@muenchow_review_2018].

Species richness is strongly associated with precipitation in the tropics [@esquivel-muelbert_biogeographic_2017;@esquivel-muelbert_seasonal_2017].
This is especially true for dry tropical ecosystems where water shortage is the main constraint to plant growth.
Still, edaphic and topographic properties are sometimes equally important in structuring vegetation of tropical ecosystems [@laurance_influence_2010; @soethe_nutrient_2008; @condit_species_2013].
Yet, the corresponding effects on the floristic composition of tropical (semi-)arid vegetation communities remain scarcely investigated [@muenchow_soil_2013; @pena-claros_soil_2012;@ulrich_climate_2014]. 
This is especially true in terms of how soil properties influence vegetation communities as soon as water is no longer the limiting factor.
For instance, nitrogen may increase primary production even in the presence of droughts and grazing pressure [@kinugasa_increasing_2012; @whitford_effects_2011].

To tackle the outlined issues at least in part, we examined how different ENSO episodes affect the floristic composition along a spatio-temporal gradient in NW Peru. 
For this, we surveyed the floristic composition on permanent plots in four years representing different ENSO episodes (dry (2011) and humid (2012) La Niña year; moderate El Niño year (2016); very humid coastal El Niño year (2017)).
Specifically, we investigated the:

(1) change in the different levels of biodiversity, namely alpha, beta, gamma and ecosystem diversity. 
The latter was represented by ordination scores and additionally visualized by means of predictive mapping.
Overall, we expected a boost in biodiversity as soon as water was no longer a limiting factor.
Regarding the spatial distribution of the floristic composition we expected a more pronounced development of vegetation assemblages along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].

(2) effects of soil characteristics and topography in shaping floristic composition and vegetation cover (biomass production). 
We expected that the influence of topographic and edaphic variables might gain in influence in wetter years.

(3) We complemented the plot survey at the landscape scale with an irrigation-fertilization experiment, thereby considering the socio-economic perspective of the ENSO variability. 
Specifically, we quantified the effect of the water-nutrient interaction on biomass development under different rainfall scenarios.
We expected that the water input of the rainfall scenario representing a Super-Niño event in combination with the fertilizer would be most conducive to biomass and biodiversity development.

@@

```{r study-area-map, echo=FALSE, fig.cap="The study area (Panel A) located in Peru (Panel B) given as a false-color composite representing the change in NDVI across the observed years. The 2011 NDVI serves as the basis for the interannual change (Panel C). Green pixels occur where the interannual difference in NDVI was greatest between 2017 and 2011. Blue color tones indicate that NDVI differences were greatest between 2012 and 2011 (red: between 2016 and 2011). Panel D: Ratio between annual precipitation and median precipitation (median Paita: 25.9mm, median Piura: 71.4mm, median Chulucanas: 272.4mm) measured at the three automated climatic stations (see section \\\\@ref(environmental-variables)). In case the precipitation was smaller than the median, we computed the ratio as: 1 / (precipitation / median(precipitation)) \\* -1. The years in which vegetation sampling took place are printed in bold (x-axis). The ENSO phases are in correspondence with the ONI index of the NOAA Climatic Prediction Center (http://origin.cpc.ncep.noaa.gov/products/analysis\\_monitoring/ensostuff/ONI\\_v5.php). We have intentionally left out the 1997/1998 Super-Niño event since this would have dominated the y-axis."}
# knitr::include_graphics(here("figures/12_map_grid.png"))
suppressPackageStartupMessages(library("raster"))
library("png")
back = as.raster(readPNG(here("figures/12_ternary_background.png")))
front = as.raster(readPNG(here("figures/12_ternary_foreground.png")))
p = readRDS(here("figures/12_map_grid.rds"))
p
```

The study area stretches for 120 km along a pronounced bioclimatic gradient in NW Peru and features a desert along the Pacific coast in the west and TDFs near the Andean foothills in the east (Fig. \@ref(fig:study-area-map)). 
The climatic gradient is at the same time a gradient of increasing human influence especially in terms of agricultural activity towards the Andean foothills [@muenchow_coupling_2013; @muenchow_woody_2013].
Agricultural activity includes field farming and extensive grazing caused by free-ranging livestock.

The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
Generally, El Niño brings more rain than usual and La Niña enforces the already dry conditions in the study area.
However, even this pattern is highly variable and can occasionally be reversed, see e.g., the years 2005, 2008 and 2012 in Fig. \@ref(fig:study-area-map)d. Additionally, so-called "local" or "coastal El Niño events" can take place as observed in 2017. 
Coastal El Niño events refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide El Niño when warm North Australian waters reach the South American coast.

@@

Therefore, to put the ecological results into context, it is important to keep in mind the varying rainfall patterns as produced by different ENSO episodes across the studied years. 
The observed years differed largely in terms of precipitation.
2011 corresponded to a dry La Niña year and the whole study area experienced drier conditions than normal.
By contrast, 2012 was a humid La Niña year, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig. \@ref(fig:study-area-map)).
2016 was a moderate El Niño year and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
2017 coincided with a coastal El Niño event and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig. \@ref(fig:study-area-map)d). 
Moreover, it is noteworthy that vegetation development changes drastically in the study area during wet episodes but remains more or less uniform during neutral years [@holmgren_extreme_2006; @muenchow_coupling_2013; @richter_monitoring_2005].

@@

We sampled 50 vegetation plots of 30 × 30 m^2^ size randomly along the climatic gradient between Paita and Chulucanas (Fig. \@ref(fig:study-area-map)).
To ensure that all parts of the study area were equally likely to be sampled, we stratified the sampling by distance to the Pacific Ocean in ten classes (see Appendix 1).
Prior to the sampling, we excluded riparian, agricultural and urban areas.
Each plot was located with a handheld Garmin GPS device in the field.
Field sampling took place towards the end of the vegetation period (March-May) in 2011, 2012, 2016 and 2017.
In each plot, we recorded all vascular plant species and determined their cover as portion of the plot area in percentage points. 
Specifically, plant cover refers to the relative projected area covered by a species [@damgaard_estimating_2014].
Prior to the analysis, we transformed the plant cover values in accordance with the decimal scale by Londo [@londo_decimal_1976].
The nomenclature follows the conventions of the Tropicos online database [@missouribotanicalgarden_tropicos_2018].

In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-nutrient experiment to quantify the beneficial effects of the water-nutrient interaction on vegetation development and biomass production.
We established the experiment in the middle of the precipitation gradient, ranging from `r rain_17[rain_17$station == "Paita", "median"] %>% pull %>% round` mm (Paita; annual median precipitation) to `r rain_17[rain_17$station == "Chulucanas", "median"] %>% pull %>% round` mm (Chulucanas; Fig. \@ref(fig:study-area-map)), on the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig. \@ref(fig:study-area-map)).
The environmental conditions are typical for the region in terms of climate, topography and soil properties (see @muenchow_soil_2013; @muenchow_woody_2013). 
Inside a sandy area devoid of woody vegetation, we placed 12 experimental plots of 3 × 3 m^2^, which were regularly spaced along a rectangular grid of 1 ha. 
Plots were assigned to one of three irrigation treatments: The first treatment represented a Super El Niño year (example of 1997/98: 1780 mm), the second represented a moderate El Niño event (example of 1991/1992: 258 mm). 
The last treatment represented the baseline with no additional water input and corresponded to the year in which the irrigation experiment was conducted (2013) which was a neutral year (54 mm; see also Fig. \@ref(fig:study-area-map)d). 
The irrigation water stemmed from rain-collecting tanks.
In addition to the water treatment, two randomly chosen plots of each irrigation treatment received a fertilization treatment of 200kg/ha granular nitrogen fertilizer (NH~4~NO~3~). 
The experiment was carried out between December 2012 and May 2013.
Plant development was monitored two times per month and included the determination of plant species [@missouribotanicalgarden_tropicos_2018] and corresponding coverage (%).
Please refer to Appendix 2 for a detailed description of the irrigation-nutrient experiment.

@@

We randomly took and mixed three soil samples from 15cm and 30cm depth in each plot in 2011. 
Soil composition should not vary too much between the years in which vegetation sampling took place since the study area largely features arenosols, i.e., a very weekly developed soil type.
On the other hand, soil chemistry can vary largely between two years.
For instance, El Niño events can increase soil respiration, i.e., soil CO\textsubscript{2} emissions, by a factor of 100, especially in the vicinity of trees [@salazar_tree_2019].
Subsequently, we measured in the laboratory variables known to be important for vegetation development: pH, electrical conductivity (µS), carbon-to-nitrogen ratio, soil texture (%), the concentrations of the cations Ca, Mg, K and Na (cmol kg^-1^), and skeletal content (%; measured as gravimetric proportion of stones \>2 mm; see [@muenchow_woody_2013] for a more detailed description).

We computed the topographic variables aspect, altitude, catchment slope, catchment area, soil wetness index, plan and vertical curvature from a digital elevation model with a 30 × 30 m^2^ resolution [@landprocessesdistributedactivearchivecenter_lp_2018].
We downloaded the normalized difference vegetation index (NDVI) raster imagery from the Moderate Resolution Imaging Spectrometer (MODIS) available for the vegetation period (March, April) for each studied year (2011, 2012, 2016, 2017) at a 250 × 250 m^2^ resolution for our study area.
Since three NDVI scenes were available for each vegetation period, we computed the mean NDVI for each vegetation period.

The interpolated precipitation values from the WorldClim data provided a spatial distribution of the average annual rainfall [@hijmans_very_2005].
Local precipitation values were collected from automated climate stations (Fig. \@ref(fig:study-area-map)d) operated by the University of Piura near Paita (1997-2018), in Piura (1991-2018) and in Chulucanas (1997-2018).

@@

We calculated four measures of biodiversity for each observed year, namely alpha, beta, gamma and ecosystem diversity [@whittaker_gradient_1967].
Alpha diversity refers to the mean number of species per plot.
Beta diversity is expressed as the range of the first Detrended Correspondence Analysis (DCA) axis (see below). 
This measure is given in units of standard deviation (SD) with four SDs corresponding to a complete species turnover.
Gamma diversity is the total number of observed species across all plots.
Ecosystem diversity was assessed with the help of a DCA [@hill_detrended_1980] to which the decimal scale transformed cover values of the species-plot matrix were subjected [@londo_decimal_1976]. 
DCA tries to condense as much as possible the observed variance of the input matrix in a low-dimensional space (usually 2-3 axes), thereby extracting the main apparent gradients.

For presenting the results of our nutrient-irrigation experiment we simply used descriptive statistics by plotting the increase in plant cover as well as water input against time for the duration of the experiment.

@@

<!-- JM: say, the nth aim in correspondence with the hypotheses made at the beginning -->
One aim of this study was to spatio-temporally predict ecosystem diversity (floristic composition) represented by ordination scores along the climatic gradient in the study area.
@muenchow_coupling_2013 have already shown that NDVI is an excellent predictor for vegetation assemblages in the study area.
Therefore, we modeled the scores of the first DCA axis representing the main observed floristic gradient in the study area with the help of a Generalized Additive Model (GAM) as a function of the interaction between NDVI and year: 
$$
\begin{aligned}
\text{Scores}_{i}\ \sim\ \mathrm{N}\left( \mu_{i},\ \sigma^{2} \right) \\
\mu_{i} = \alpha + f_{\text{year}_{i}}\left( \text{NDVI}_{i} \right) + \beta_{1}\text{year}_{i}
\end{aligned}
$$

@@

where *i* refers to the *i*th observation, *$\alpha$* is the intercept, *$\beta$*~1~ is the estimated coefficient for year and *f* is a spline smoother for the NDVI:year interaction.

>>>>

title: "Monitoring and predictive mapping of floristic biodiversity along a climatic gradient in ENSO's terrestrial core region, NW Peru"
bibliography: ../transect_refs.bib
csl: ../ecography.csl
# biblio-style: "natbib"
link-citations: true
---

```{r setup, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("here")

# needed for dates in experiment figure species richness, otw months are displayed in German
Sys.setlocale("LC_TIME", "C")

# make sure that figures are allowed to float (hold the assigned position and are not always put on top of a page)
knitr::opts_chunk$set(fig.pos = 'H')
```

@@

Studying the relationship between environmental gradients and species turnover has become a cornerstone of biogeography and community ecology since Humbold's and Bonpland's travels to South America more than 200 years ago [@macarthur_geographical_1984; @whittaker_gradient_1967]. 
Naturally, water availability drives floristic composition and biomass production in (semi-)arid regions such as NW Peru [@muenchow_coupling_2013; @muenchow_woody_2013], which is the area under study in this manuscript.
In fact, water availability is probably the most important limiting abiotic factor in NW Peru's tropical dry forests, where it controls plant recruitment, establishment and survival, seedling production, vegetation cover as well as species diversity [@balvanera_distribution_2011;@espinosa_what_2011;@salazar_tree_2020].
However, water availability does not only vary in space but also in time, especially in regions heavily affected by the El Niño Southern Oscillation (ENSO).
<!--
A super-humid El Niño (EN) episode benefits other species than a super-dry and fire-bringing La Niña (LN) episode [@richter_monitoring_2005].
-->

ENSO is a recurrent climate phenomenon causing anomalies around the globe [@capotondi_understanding_2015].
ENSO consists of alternating episodes.
El Niño represents the warm and La Niña the cold episode of this cycle. 
ENSO neutral refers to periods when neither El Niño nor La Niña episodes are present. 
The study area is part of the core region where the ENSO phenomenon exerts its greatest terrestrial influence [@richter_monitoring_2005]. 
Generally, El Niño episodes bring more rain than usual and La Niña episodes enforce the already dry conditions in the study area [<50 mm of annual precipitation along the coast of NW Peru; Fig. \@ref(fig:study-area-map)D;  @rollenbeck_climatic_2015].

Especially El Niño episodes provoke drastic ecological responses [@Kogan.2017] but occur irregularly (five times over the last 20 years in the study area), and are highly variable in terms of intensity.
<!--
The anomalies show recurrent patterns to a certain degree but are still highly variable especially on a local scale. 
For instance, in 2016 the predicted Super-Niño rainfalls largely failed to materialize along the coasts of Ecuador and Peru.
Just a year later, unexpected, torrential rainfalls hit the same coasts during a supposedly ENSO neutral year with a magnitude comparable to that of the Super-Niño event in 1997/98. Originating from a so-called coastal EN, the 2017 rainfalls came as a complete surprise, and caused the unprepared population unbearable hardship [@fraser_peru_2017; @ramirez_understanding_2017].
-->
Nevertheless, the eco-climatic impact of La Niña episodes is comparable in magnitude with that of El Niño episodes.
For instance, La Niña episodes were most likely the main driver for the hiatus in global warming at the beginning of the 21^st^ century [@kosaka_recent_2013].
Therefore, one objective of this manuscript is to study the effect of different ENSO episodes, including La Niña episodes, on the floristic composition, comprised of all observed vascular plants, of the same ecosystem.
In any case, ENSO drives a large amount of the interannual variability of woody plant growth across the tropics [@rifai_enso_2018] though the specific response to El Niño episodes may vary from region to region [@phillips_changes_2009].
Overall, the effect of  extreme events is even more pronounced in (semi)-arid ecosystems [@holmgren_extreme_2006], and is especially true for ENSO's terrestrial core region in NW Peru and SW Ecuador [@muenchow_woody_2013; @rollenbeck_climatic_2015].

NW Peru is also home to the tropical dry forest (TDF) ecosystems.
TDFs are characterized by a high rate of endemism and feature an outstanding biodiversity which is almost as high as that of humid tropical forests [@espinosa_what_2011;@pennington_tropical_2018]. 
At the same time, they are the most threatened tropical ecosystem [@miles_global_2006], and in general chronically understudied [@escribano-avila_biodiversity_2017;@muenchow_review_2018].
Despite their unique biodiversity and threatened status, the TDFs of NW Peru are vastly understudied, and the spatio-temporal effect of extreme events like El Niño or La Niña episodes on the biodiversity patterns in this ecosystem remains largely unknown.

Statistical learning has a long tradition in biodiversity research for detecting patterns in ecological data and predicting from these patterns [@guisan_predictive_2000].
It is especially useful for quantifying species-environment relationships and for predicting the occurrence of species or species richness (predictive mapping), and therefore is an indispensable instrument for biodiversity management [@fremout_mapping_2020].
Furthermore, predictive mapping can reveal ecological changes in space and time which otherwise might be hard or impossible to detect [@lovelace_geocomputation_2019a].
Therefore, it is a crucial tool for gaining detailed insights into ecosystem dynamics, which other more descriptive methods such as global biodiversity measures and ordination techniques cannot capture.
We aim at spatially predicting, for the first time, how different ENSO episodes might change and reshape the spatial distribution of the floristic composition in the study area.

Species richness is strongly associated with precipitation in the tropics [@esquivel-muelbert_biogeographic_2017;@esquivel-muelbert_seasonal_2017].
This is especially true for dry tropical ecosystems where water shortage is the main constraint to plant growth.
Still, edaphic and topographic properties are sometimes equally important in structuring vegetation of tropical ecosystems [@laurance_influence_2010; @soethe_nutrient_2008; @condit_species_2013].
Yet, the corresponding effects on the floristic composition of tropical (semi-)arid vegetation communities remain scarcely investigated [@muenchow_soil_2013; @pena-claros_soil_2012;@ulrich_climate_2014]. 
This is especially true in terms of how soil properties influence vegetation communities as soon as water is no longer the limiting factor.
For instance, nitrogen may increase primary production even in the presence of droughts and grazing pressure [@kinugasa_increasing_2012; @whitford_effects_2011].
Hence, this study investigates how much the impact of edaphic and topographic variables varies in explaining the floristic composition between dry and humid ENSO years.
Additionally, an irrigation-fertilization experiment should help to quantify in more detail the beneficial effects of the water-nutrient interaction on vegetation diversity and cover.

In summary, in this manuscript we study the effects of different ENSO episodes on biodiversity patterns and floristic composition  as well as the effect of environmental variables on the floristic composition under varying rainfall patterns.
Specifically, we tested the following three hypotheses:

(1) We investigated the change in the different levels of biodiversity along a pronounced climatic gradient across different ENSO episodes. 
Overall, we expected a boost in biodiversity as soon as water was no longer the predominant limiting factor.
We were particularly interested in how different ENSO episodes modify the spatial distribution of the floristic composition.
Overall, we expected a more pronounced development of vegetation formations along the climatic gradient with wetter conditions [@engelbrecht_drought_2007].

(2) Water availability limits plant cover and species richness along the entire climatic gradient in the (semi-)arid study area.
Therefore, we were interested in evaluating the effects of soil characteristics and topography changed in shaping floristic composition and vegetation cover (biomass production) with varying rainfall patterns as caused by different ENSO episodes. 
We expected that the influence of topographic and edaphic variables might gain in influence in wetter years, i.e., as soon as water is no longer the limiting factor.

(3) To quantify the beneficial effects of the water-nutrient interaction on vegetation cover development and species richness, we executed an irrigation-fertilization experiment simulating different ENSO rainfall scenarios.
We expected that the water input of the rainfall scenario representing a Super-Niño episode in combination with the fertilizer would be most conducive to plant cover and species richness development.

@@

```{r study-area-map, echo=FALSE, fig.cap="The study area (Panel A) located in Peru (Panel B) given as a false-color composite representing the change in NDVI across the observed years. The 2011 NDVI serves as the basis for the interannual change (Panel C). Green pixels occur where the interannual difference in NDVI was greatest between 2017 and 2011. Blue color tones indicate that NDVI differences were greatest between 2012 and 2011 (red: between 2016 and 2011). Panel D: Ratio between annual precipitation and median precipitation (median Paita: 25.9mm, median Piura: 71.4mm, median Chulucanas: 272.4mm) measured at the three automated climatic stations (see section \\@ref(environmental-variables)). In case the precipitation was smaller than the median, we computed the ratio as: 1 / (precipitation / median(precipitation)) \\* -1. The years in which vegetation sampling took place are printed in bold (x-axis). The ENSO phases are in correspondence with the ONI index of the NOAA Climatic Prediction Center (http://origin.cpc.ncep.noaa.gov/products/analysis\\_monitoring/ensostuff/ONI\\_v5.php). We have intentionally left out the 1997/1998 Super-Niño episode since this would have dominated the y-axis. Abbreviations: EN = El Niño episode; LN = La Niña episode."}
# knitr::include_graphics(here("figures/12_map_grid.png"))
suppressPackageStartupMessages(library("raster"))
library("png")
back = as.raster(readPNG(here("figures/12_ternary_background.png")))
front = as.raster(readPNG(here("figures/12_ternary_foreground.png")))
p = readRDS(here("figures/12_map_grid.rds"))
p
```

The study area stretches for 120 km along a pronounced bioclimatic gradient in NW Peru and features a desert along the Pacific coast in the west and TDFs near the Andean foothills in the east (Fig. \@ref(fig:study-area-map)). 
The climatic gradient is at the same time a gradient of increasing human influence especially in terms of agricultural activity towards the Andean foothills [@muenchow_coupling_2013; @muenchow_woody_2013].
Agricultural activity includes cultivation of annual crops and fruit trees and extensive grazing caused by free-ranging livestock.

The ENSO phenomenon heavily impacts the study area [@richter_monitoring_2005]. 
In general, El Niño episodes are associated with more rain than usual while La Niña episodes amplify the already dry conditions.
However, even this pattern is highly variable and can occasionally be reversed (see e.g., the years 2005, 2008 and 2012 in Fig. \@ref(fig:study-area-map)d). 
Additionally, so-called "local" or "coastal El Niño episodes" can take place as observed in 2017. 
Coastal El Niño episodes refer to the regional warming of coastal waters in the Gulf of Guayaquil in December as opposed to the Pacific basin-wide El Niño episode, which occurs when warm North Australian waters reach the South American coast.

@@

Therefore, to put the ecological results into context, it is important to keep in mind the varying rainfall patterns as produced by different ENSO episodes across the studied years. 
The observed years differed largely in terms of precipitation.
2011 corresponded to a dry La Niña episode and the whole study area experienced drier conditions than normal.
By contrast, 2012 was a humid La Niña episode, and 1.5 times and 3 times more rain fell in Piura and Chulucanas, i.e., in the middle to eastern part of the study area (Fig. \@ref(fig:study-area-map)A).
2016 was a moderate El Niño episode and rainfall was double the median input in Paita and Piura, whereas in Chulucanas the median value was almost reached. 
2017 coincided with a coastal El Niño episode and is the only studied year during which the whole study area experienced exceptionally wet conditions: Paita received almost `r rain_17[rain_17$station == "Paita", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Paita", "precip"] %>% pull %>% round` mm), Piura almost `r rain_17[rain_17$station == "Piura", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Piura", "precip"] %>% pull %>% round` mm) and Chulucanas `r rain_17[rain_17$station == "Chulucanas", "dev"]  %>% pull %>% round` times (`r rain_17[rain_17$station == "Chulucanas", "precip"] %>% pull %>% round` mm) more rain than on median (Fig. \@ref(fig:study-area-map)D). 
Moreover, it is noteworthy that vegetation development changes drastically in the study area during wet episodes but remains more or less uniform during neutral years [@holmgren_extreme_2006; @muenchow_coupling_2013; @richter_monitoring_2005].

@@

We sampled 50 vegetation plots of 30 × 30 m^2^ size along the climatic gradient between Paita and Chulucanas (Fig. \@ref(fig:study-area-map)).
To ensure that all parts of the study area were equally likely to be sampled, we stratified the random sampling by distance to the Pacific Ocean in ten classes (see Appendix 1).
Prior to the sampling, we excluded riparian, agricultural and urban areas.
Each plot was located with a handheld Garmin GPS device in the field.
Field sampling took place towards the end of the vegetation period (March-May) in 2011, 2012, 2016 and 2017.
In each plot, we recorded all vascular plant species and determined their cover as portion of the plot area in percentage points. 
Specifically, plant cover refers to the relative projected area covered by a species [@damgaard_estimating_2014], and is a proxy for biomass.
Prior to the analysis, we transformed the plant cover values in accordance with the decimal scale by Londo [@londo_decimal_1976].
The nomenclature follows the conventions of the Tropicos online database [@missouribotanicalgarden_tropicos_2018].

In addition to the plot survey at the landscape scale, we conducted a small-scale irrigation-fertilization experiment to quantify the beneficial effects of the water-nutrient interaction on species richness, plant cover and biomass production.
We established the experiment in the middle of the study area on on a protected dryland forest inside the campus of the University of Piura (5°10'S, 80°38'W, see Piura in Fig. \@ref(fig:study-area-map)A).
Hence, the experiment was located in the same ecosystem as the landscape study (see previous paragraph) and more specifically in the middle of the observed precipitation gradient, ranging from `r rain_17[rain_17$station == "Paita", "median"] %>% pull %>% round` mm (Paita; annual median precipitation) to `r rain_17[rain_17$station == "Chulucanas", "median"] %>% pull %>% round` mm (Chulucanas; Fig. \@ref(fig:study-area-map)A).
The environmental conditions are typical for the region in terms of climate, topography and soil properties (see @muenchow_soil_2013; @muenchow_woody_2013). 
Inside a sandy area devoid of woody vegetation, we placed 12 experimental plots of 3 × 3 m^2^, which were regularly spaced along a rectangular grid of 1 ha. 
Plots were assigned to one of three irrigation treatments: The first treatment represented a Super El Niño episode (example of 1997/98: 1780 mm), the second represented a moderate El Niño episode (example of 1991/1992: 258 mm). 
The last treatment represented the baseline with no additional water input and corresponded to the year in which the irrigation experiment was conducted (2013) which was a neutral episode (54 mm; see also Fig. \@ref(fig:study-area-map)D). 
The irrigation water stemmed from rain-collecting tanks.
In addition to the water treatment, two randomly chosen plots of each irrigation treatment received a fertilization treatment of 200kg/ha granular nitrogen fertilizer (NH~4~NO~3~). 
The experiment was carried out between December 2012 and May 2013.
Monitoring was executed two times per month and included the identification of plant species [@missouribotanicalgarden_tropicos_2018] and corresponding coverage (%).
Please refer to Appendix 2 for a detailed description of the irrigation-fertilization experiment.

@@

We randomly took and mixed three soil samples from 15cm and 30cm depth in each plot in 2011. 
Soil composition should not vary too much between the years in which vegetation sampling took place since the study area largely features arenosols, i.e., a very weekly developed soil type.
On the other hand, soil chemistry can vary largely between two years.
For instance, El Niño episodes can increase soil respiration, i.e., soil CO\textsubscript{2} emissions, by a factor of 100, especially in the vicinity of trees [@salazar_tree_2020].
Subsequently, we measured in the laboratory variables known to be important for vegetation development: pH, electrical conductivity (µS), carbon-to-nitrogen ratio, soil texture (%), the concentrations of the cations Ca, Mg, K and Na (cmol kg^-1^), and skeletal content (%; measured as gravimetric proportion of stones \>2 mm; see @muenchow_woody_2013 for a more detailed description).

We computed the topographic variables aspect, altitude, catchment slope, catchment area, soil wetness index, plan and vertical curvature from a digital elevation model with a 30 × 30 m^2^ resolution [@landprocessesdistributedactivearchivecenter_lp_2018].
We downloaded the normalized difference vegetation index (NDVI) raster imagery from the Moderate Resolution Imaging Spectrometer (MODIS; @landprocessesdistributedactivearchivecenter_lp_2018) available for the vegetation period (March, April) for each studied year (2011, 2012, 2016, 2017) at a 250 × 250 m^2^ resolution for our study area.
Since three NDVI scenes were available for each vegetation period, we computed the mean NDVI for each vegetation period.

Local precipitation values were collected from automated climate stations (Fig. \@ref(fig:study-area-map)d) operated by the University of Piura near Paita (1997-2018), in Piura (1991-2018) and in Chulucanas (1997-2018).

@@

We calculated three measures of biodiversity for each observed year, namely alpha, beta and gamma diversity [@whittaker_gradient_1967].
Alpha diversity refers to the mean number of species per plot.
Gamma diversity is the total number of observed species across all plots.
Beta diversity, in the sense of species turnover, is expressed as the range of the first Detrended Correspondence Analysis (DCA) axis.  
The unit of a DCA axis is given in units of standard deviation (SD).
4 SDs correspond to a complete species turnover while a difference between 1 to 1.4 SD units already correspond to half a change in species composition [@legendre_numerical_2012]. 
Moreover, the floristic composition was assessed with the help of a DCA [@hill_detrended_1980] to which the decimal scale transformed cover values of the species-plot matrix were subjected [@londo_decimal_1976]. 
Floristic composition implicitly represents the named biodiversity measures, and additionally considers plant cover. DCA tries to condense as much as possible the observed variance of the input matrix in a low-dimensional space (usually 2-3 axes), thereby extracting the main apparent gradients.
To test if beta diversity and floristic composition, respectively, varied significantly between years, we analyzed the multivariate homogeneity of group dispersions [@anderson_multivariate_2006].

@@

One objective of this study was the spatio-temporal mapping of floristic composition represented by the scores of the first DCA axis along the climatic gradient in the study area.
Please note that the resulting prediction map not only shows species composition but also differences in beta diversity on the predicted pixel level.
This is because the units of a DCA axis are given in SDs (see also previous section).

@muenchow_coupling_2013 have already shown that NDVI is an excellent predictor for vegetation formations in the study area.
Therefore, we modeled the scores of the first DCA axis representing the main observed floristic gradient in the study area with the help of a Generalized Additive Model (GAM) as a function of the interaction between NDVI and year:

$$
\begin{aligned}
\text{Scores}_{i}\ \sim\ \mathrm{N}\left( \mu_{i},\ \sigma^{2} \right) \\
\mu_{i} = \alpha + f_{\text{year}_{i}}\left( \text{NDVI}_{i} \right) + \beta_{1}\text{year}_{i}
\end{aligned}
$$

@@

where *i* refers to the *i*th observation, *$\alpha$* is the intercept, *$\beta$*~1~ is the estimated coefficient for year and *f* is a spline smoother for the NDVI:year interaction. 
Please refer to Appendix 3 for a visualization of the non-linear relationship between the response variable (ordination scores) and NDVI by year.

The sole purpose of the model is the spatial prediction of the ordination scores, not statistical inference or the interpretation of the models' coefficients.
We will use variation partitioning later on to assess how the influence of environmental predictors has changed over the studied years (see section \@ref(impact-of-environmental-variables)). 
Consequently, here we are only interested in a good predictive performance of the model to make sure that it is able to generalize, i.e., that it is able to predict reasonably well unseen data [@james_introduction_2013; @kuhn_applied_2013]. 
To make sure this is the case, we used spatial cross-validation.
Overall, cross-validation accounts for over-optimistic predictions on the training set by randomly splitting a data set into partitions used for training and testing [@brenning_spatial_2012;@lovelace_geocomputation_2019a]. 
Spatial cross-validation is an extension that measures a model's ability to spatially predict the response variable in the presence of spatial autocorrelation [@schratz_hyperparameter_2019]. 
We ran a 100-repeated 5-fold spatial cross-validation in which the partitioning is based on *k*-means clustering (*k* = 5; @russ_data_2010).
We also made sure that all annual observations of one plot were jointly placed in either the test or the training data set [@meyer_improving_2018]. 
We used the normalized root mean-square error (NRMSE) in percent as performance measure for the spatial cross-validation.
